Icicle: write once, run once

### 2016-05-22 2016fhpc/section/01-Introduction.tex
Evaluation rules

The contributions of this paper are:
\begin{itemize}
\item
we describe restrictions on Icicle which guarantee any two queries on the same table can be fused together\sref{s:IcicleSource};
\item
we show an intermediate fold-based language, which makes fusion a simple matter of appending two programs\sref{s:IcicleCore};
\item
and we present benchmarks of our generated code, which compares favourably against standard unix utilities and outperforms our existing system by orders of magnitude\sref{s:Benchmarks}.
\end{itemize}

>>>>

The contributions of this paper are:
\begin{itemize}
\item
we show a typesystem that guarantees any two queries on the same table can be fused together, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};

\item
we show an intermediate fold-based language, which makes fusion a simple matter of appending two programs\sref{s:IcicleCore};

\item
and we present benchmarks of our generated code, which compares favourably against standard unix utilities and outperforms our existing system by orders of magnitude\sref{s:Benchmarks}.
\end{itemize}

### 2016-05-22 2016fhpc/section/02-Source.tex
Evaluation rules

\TODO{update this to fit new grammar}
Expressions can be variable names, primitive operators such as addition and division, function application, or nested queries.
Function application @x Exp*@ is the name of a top-level function applied to any number of arguments.
Primitive application @Prim Exp*@ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.

Queries can define folds over the input, let-bindings, filtering according to some predicate, or grouping by some key.
The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
Count can be expressed by the following fold:
\begin{code}
fold count = 0 then count + 1
\end{code}

@@

@@

Group takes a key to group by, and a query to perform on the values of each grouping.
This query groups by the company code, and counts the number of records in each company.
\begin{code}
group company in count
\end{code}

@Fun@ is used for function definitions.
The @count@ function takes no arguments and returns a fold counting the number of elements.
\begin{code}
function count
 = fold c = 0 then c + 1
\end{code}

Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
It returns a fold that starts at zero and for every element, increases the old sum by @value@.
\begin{code}
function sum (value : Element Int)
 = fold s = 0 then s + value
\end{code}

The definition of mean is just the sum divided by the count.
\begin{code}
function mean (value : Element Int)
 = sum value / count
\end{code}

@@

Some examples with their types:
\begin{code}
1             :         Int
open          : Element Int
open > 1      : Element Int
sum           : Element Int -> Aggregate Int
sum open      : Aggregate Int
sum open > 1  : Aggregate Bool
\end{code}

Folds have a name and two expressions.
The first expression, the initialiser, is pure.
The second expression, the update, has the fold's name bound to an @Element@ of the previous value.
The type of the expression must be @Element@.
The final return of the fold is an @Aggregate@.
\begin{code}
fold x = T then Element T : Aggregate T
\end{code}

Let-bindings are independent of the modality.

For filters, the predicate is an @Element@, which could be thought of as a stream of booleans.
The rest of the query must be an @Aggregate@; that is, it must return a scalar rather than a stream, to enforce the bounded buffer restriction explained in \sref{s:IcicleSource:bounded}.
\begin{code}
filter (Element Bool) in (Aggregate T)
\end{code}

Grouping is similar to filters; the key must be an @Element@, while the group value must be an @Aggregate@.
\begin{code}
group (Element Key) in (Aggregate Value)
\end{code}

The following program attemps to find the number of entries where the open price is above the mean of the close price.
This requires two passes over the data, as we do not know what the mean of all the data is until we have seen all the data.
This program violates the single-pass restriction, and is outlawed by the typesystem.
The left side of the @>@ has type @Element Int@, while the right side has type @Aggregate Int@, and these two are mutually exclusive.
\begin{code}
filter open > mean close
in count
(Error: (>) must be Element and Aggregate)
\end{code}

Here is an example to highlight one benefit of tracking these modalities.
Suppose we have a table with two fields: key and value.
We wish to find the mean of values whose key matches the most recent key.
Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only thosewith the right key.

@@

>>>>

Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.

@@

The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
Count can be expressed by the following fold:
\begin{code}
fold count = 0 then count + 1
\end{code}

@@

Group takes a key to group by, and a query to perform on the values of each grouping.
This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
\begin{code}
group ((open - close) / 100) in count
\end{code}

$\mi{Fun}$ is used for function definitions.
The @count@ function takes no arguments and returns a fold counting the number of elements.
\begin{code}
function count
 = fold c = 0 then c + 1;
\end{code}

Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
It returns a fold that starts at zero and for every element, increases the old sum by @value@.
\begin{code}
function sum (value : Element Int)
 = fold s = 0 then s + value;
\end{code}

The definition of mean is just the sum divided by the count.
\begin{code}
function mean (value : Element Int)
 = sum value / count;
\end{code}

$\mi{Query}$ defines an aggregate query which must return a single value, and gives it a name:
\begin{code}
query count_all_records = count;
\end{code}

Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
All queries here operate over the same input table, and will be fused into a single loop over the input data.

@@

Here is an example to highlight one benefit of tracking these modalities.
Suppose we have a table with two fields: key and value.
We wish to find the mean of values whose key matches the most recent key.
Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only those with the right key.

@@

\subsection{Type system}

The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{T_M}$ means that under context $\Gamma$, the expression has a given value type and modality.
$\TypecheckP{\mi{Prim}}{T_\to}$ associates a primitive with its function type.
$\TypecheckApp{T_\to}{\ov{T_M}}{T_M}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
The next two judgments,
$\TypecheckS{\Gamma}{\mi{Fun}}{\Gamma}$ and
$\TypecheckS{\Gamma}{\mi{Exp}}{\Gamma}$
both takes an input environment and function or query, and returns an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.

The rule (TcVar) performs variable lookup in the context, while (TcNat) shows that number constants have type $\NN$.
(TcBox) shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.

(TcPrimApp) and (TcFunApp) find the result type of primitive and function application.
They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{T_\to}{\ov{T_M}}{T_M}$ to find the return type and modality of the application.
The set-comprehension syntax $\{e_i\}_i$ means that this rule allows any number of arguments.

(TcLet) is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.

(TcFold) finds the type of a @fold@ expression.
The initial value of the fold must be pure.
The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
The final return type of the fold is an @Aggregate@.

(TcFilter) requires its predicate to be a stream of booleans ($@Element@~\BB$).
The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
(TcGroup) is similar to filter, except returning a @Map@.

The rules (PrimArith), (PrimRel) and (PrimLookup) give function types for each primitive.

The next two rules show how function application boxes and unboxes modalities.

(AppArgs) applies when a function is applied to exactly the right arguments, and simply returns the result of the function.

(AppRebox) is used when a pure function is applied to a non-pure value.
If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
Note that (TcBox) rule can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.

(CheckFun) inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.

(CheckQuery) checks a query's types, and makes sure it returns an @Aggregate@.

Finally, (CheckTop) checks a whole top-level program.
It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
Then, each query is checked, with the table columns and all functions available.
The final returned context is all query names and their types.

\subsection{Evaluation}

The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.

We define simple values $V$ as numbers, booleans, maps of values, or tuples of values.
Next, stream values $V'$ correspond to modalities: pure computations are represented as a single value;
@Element@s as a stream transformer from a heap to a value;
and @Aggregate@s as folds defining an initial state and update and extract functions.

The evaluation rules operate over a slightly modified expression type, $E'$, which is an $\mi{Exp}$ with $V'$ as values instead of just simple values and natural numbers.
This is to simplify typechecking over $V'$, and as $V'$ only appear in restricted form for evaluation.
We also assume that all functions have been inlined before evaluation.

The judgment $\SourceStepX{E'}{V'}$ defines a big-step evaluation relation.

(EVal) applies when the expression is alread a stream value.
This corresponds, roughly, to the (TcNat) typechecking rule.

(EBoxStream) and (EBoxFold) convert a pure value to a stream and fold, respectively.
The stream transformer ignores the input stream and returns the pure value.
The fold has a unit accumulator, and the extract function simply returns the pure value.

Next (EPrimValue), (EPrimStream) and (EPrimFold) evaluate primitive applications.
For pure values, the values are unboxed and the primitive is applied.
For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
Folds are the most complicated: the fold accumulators are tupled together, for example $(\times_i~z_i)$ being the tuple of all the initial accumulators.
At each step, the accumulators for all the input folds are updated with their own update function, and the input heap.
Finally, the extract function extracts the final values of the input folds, and applies the primitive.

The (ELet) rule is fairly standard, evaluating the bound to a value, substituting it in, and evaluating the rest.

(EFold) is one of the most complex rules, as it introduces recursion to the streams.
It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
In the update expression, the fold binding $x$ is substituted with a stream that looks up $x$ in its input heap.
This action-at-a-distance is necessary because the value of $x$ depends on the value of the fold, which is not known yet.
Finally, the return fold is constructed, which binds $x$ to the current fold accumulator.
The extract function for this is the identity, simply returning the accumulator.

(EFilter) evaluates its predicate to a stream of booleans, and the rest to a fold.
The result is a fold which only applies the update function when the predicate stream is true.

(EGroup) is rather similar to filter, except that it returns a map.
Here, we start with an empty map of bottoms.
At every update step, we find the key and look up the current accumulator for that key in the map, or the fold initialiser if it is bottom.
We then apply the update function to the accumulator, and update that key in the map.
For the extract function, we simply apply the extract to each element of the map.

In order to execute these evaluation rules on an actual input table, a few steps are necessary.
First, variables referring to column names such as @open@ or @close@ must have stream values substituted in, so that the actual value will be looked up in the input heap.
For example, $e[@open@~:=~@Stream@~(\lam{\sigma} \sigma~@open@)]$.
Next, the query must be converted to a @Fold@.
Then, for each record, the update function would be called with an input heap binding the record values: $\{@open@ = 1, @close@ = 2\}$ and so on.
This occurs for every record in the table.
Finally, the extract function is called, which gives the result of the query.
